Template for indicator documentation

### 2023-09-11 functional_plant_indicators_seminat.Rmd
RenderBook and remove sysDate. Fixed some typos in gjengoing and image widths

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(readxl)
library(sf)
library(plyr)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019, contains semi-natural grasslands(T32), semi-natural wet meadows (V10), heathlands (T34), and we also decided to include near-semi-natural meadows (T41) ) and (2) from the ASO monitoring scheme (cf. Bär et al. 2021, targets semi-natural grasslands (T32) only)
- as reference data we use (3) generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- (4) Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the analysis in this document, we only use the plots which were classified as one of the semi-natural ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) ASO monitoring data
ASO stands for 'arealrepresentativ overvåking av semi-naturlig eng', i.e. 'area representative monitoring of semi-natural meadows'. In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape. In the future, all ASO data will hopefully be available in Miljødirektoratets kartkatalog.

@@

@@

```
The figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests. According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.

>>>>

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(readxl)
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

- as test data we use plant community data (1) from the ANO monitoring scheme (cf. Tingstad et al. 2019) and (2) from the ASO monitoring scheme (cf. Bär et al. 2021)
- as reference data we use (3) generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- (4) Swedish plant indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance from Tyler et al. (2021)

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 18 evenly distributed vegetation analyses in 1 x 1 m plots (up to 18, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the analysis in this document, we only use the plots which were classified as one of the semi-natural ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) ASO monitoring data
ASO stands for 'arealrepresentativ overvåking av semi-naturlig eng', i.e. 'area representative monitoring of semi-natural meadows'. In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape yet. In the future, all ASO data will probably be available through Miljødirektoratets kartkatalog.

@@

```
In the ANO data, deviations in the grazing/mowing index occur mainly in semi-natural grasslands (T32) on intermediate soils in terms of limestone (C3, C4, C6, C5 [but the latter is one category higher in limestone than intermediate]). The deviations towards low grazing/mowing values indicate that management and use are too extensive.

@@

```
In the ASO data, the nitrogen deviations indicate too high nitrogen content in limestone richer (intermediate to very rich) semi-natural grasslands.

### 2023-09-11 functional_plant_indicators_wetland.Rmd

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 18 evenly distributed vegetation analyses in 1 x 1 m plots (up to 18, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which as of today comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

>>>>

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

### 2023-09-11 rosslyng.Rmd

---
title: "DEPRECATED: Areal uten død eller skadet røsslyng i kystlynghei"
output:
  html_document
---

# DEPRECATED: Areal uten død eller skadet røsslyng i kystlynghei

@@

_Author and date:_
Anders L. Kolstad

```{r, echo=F}
Sys.Date()
```

@@

## Introduction
Here I quickly go through the ANO data set for the years 2019-2021 to investigate the frequeny of poins that fall within coastal heathands (kystlynghei). This will decide wether we can use this data to inform an indikcator about the proportion of dead Calluna vulgaris (røsslyng).

@@

Imnport and list terms: 
```{r}
ano <- sf::st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                   layer = "ANO_SurveyPoint")
#st_crs(ano)
names(ano)

@@

```{r}
table(ano$hovedtype_250m2)
```
Only 26 of the 250m2 circles are kystlynghei.

## Conclusion
Because of the random allocation of sampling points, not enough points fall in coastal heathlands to allow us to create a trustworth indicator for this nature type. With a few more years of data we might end up with about 100 points, which is borderline what we need in order to calculate condition values at a regional level with some level of precition.

>>>>

# (PART\*) DEPRECATED{-}

# Areal uten død eller skadet røsslyng i kystlynghei

@@

_Author and date:_

Anders L. Kolstad

March 2023

@@

## Introduction
Here I quickly go through the ANO data set for the years 2019-2021 to investigate the frequeny of poins that fall within coastal heathlands (kystlynghei). This will decide whether we can use this data to inform an indicator about the proportion of dead _Calluna vulgaris_ (røsslyng).

@@

Import and list terms: 
```{r}
ano <- sf::st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                   layer = "ANO_SurveyPoint")
#st_crs(ano)
names(ano)

@@

```{r}
table(ano$hovedtype_250m2)
```
Only 26 of the 250m2 circles are kystlynghei. The variable of interest is called _Dekning av død/skadet røsslyng_ and is only recorded on these 26 circles.

## Conclusion
Because of the random allocation of sampling points, not enough points fall in coastal heathlands to allow us to create a trustworthy indicator for this nature type. With a few more years of data we might end up with about 100 points, which is borderline what we need in order to calculate condition values at a regional level with some level of precision.
